Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2021-01-08 thesis/thesis.tex
thesis: Review State of the Art

An initial approach for finding OCs is the search of overdensities along the galactic disk. In general, this is the starting point but,
although it seems simple, it presents a fundamental problem already discussed. The near field around the OC is filled with two types of
distinct star populations: those who belong to the OC (tens or hundreds to a few thousand) and a background made up of thousands or
millions of stars that are not. Finding out which stars belong to the first group is the problem faced in this work.
This selection is crucial to properly characterize the fundamental properties of the cluster
(dynamics, total mass, age, chemical composition, etc.).

@@

However, as shown in Figures \ref{fig:pm_ngc_6494} and \ref{fig:parallax_ngc_6494}, in general it is not as easy and
becomes necessary to consider other parameters such as distances, or even metallicity and age (derived from isochrone curves)
and requires photometric data for the stars within the studied field.

@@

At this point, it seems reasonable that the study of each individual group requires its own parametrization and technique
to identify it. With Gaia DR2, which contains great high-quality astrometric and photometric data for a huge number of stars,
a new opportunity arises to simplify and optimize these processes. In this context, new approaches are being developed
to improve current detection methods and automate OC characterization.

@@

In this work we make use of Gaia DR2 since, although Release 3 (Gaia DR3) is imminent, it has not been available in time for us.
Gaia DR2 is a multidimensional dataset obtained by ESA's Gaia mission (located at L2, 1.5 million kilometers from Earth) and
operational since 2014. The catalogue has high precision and accuracy astrometric data for more than 1.7 billion stellar sources,
and magnitudes in three photometric filters (G, BP and RP) for more than 1,300 million sources.
80\% of Gaia DR2 sources are weaker than $G_{mag} \approx 18$. For magnitudes $G \approx 21$ nominal uncertainties reach
$2 mas$ for parallaxes and $5 mas \cdot years^{-1}$ in proper motions. But for closer sources, (at the bright end, $G < 14$),
the precision is of the order of $0.02 mas$ and $0.05 mas \cdot year^{-1}$ respectively \cite{cantat2018gaia}. In general,
open clusters, which is the main subject of this work, are at optimal data thresholds.

In this context, and taking advantage of the opportunity provided by Gaia DR2, new approaches are being developed to improve current
detection methods and automating OC characterization as much as possible. This is the case of tools such as Clusterix 2.0
\cite{balaguer2020clusterix}, TOPCAT \cite{taylor2005topcat}, and VOSA \cite{bayo2008vosa}, (all of them developed by the Virtual Observatory).
These tools work together and offer good results, although it is not possible to use them in an unsupervised process and free of certain
different parametrization in each case.

@@

TOPCAT alone does not offer an independent discrimination mechanisms to researchers.
They should try an initial selection, starting from the map of proper motions,
by delimiting the area where the cluster presumably appears. When this is not possible, it is necessary to have a
certain knowledge about the cluster profile and to parametrize some approximated values.
In both cases, the definition of the selected set must then be optimized step by step until reaching a result that can be considered reasonable.

Another tool is Clusterix 2.0, which is an interactive web-based tool \cite{balaguer2020clusterix}.
It takes the proper motion diagram without making any prior assumptions about the membership of the candidate star
and determines empirically the frequency functions.

@@

$(a, b)$ are referred to the proper motion configuration space, while a point located in the center of the cell $(i, j)$ provides the maximum
contribution for calculating the local density. In this context, the number $h$ is called the \emph{smoothing} parameter and it is measured in
the same units as proper motion.

With this same purpose of applying non-parametric methods, various clustering algorithms have been applied in the past, UPMASK \cite{krone2014upmask}
was developed to only use photometry and positions, and then it was adapted to the Gaia data \cite{cantat2018gaia} based only on proper motion and
parallax. A similar approach has been carried out using DBSCAN and machine learning algorithms to discover new clusters.

In this sense, Clusterix also claims to be a non-parametric method, but it critically depends on the initial selection for the field sizes to be analyzed.

@@

This fact implies defining three areas or regions with different radius. The first 'c+f' corresponds to the one in which the cluster members are
presumed to be contained together with other star fields that are not part of the cluster. The second, broader, assumes that it only contains
stars in an extended visual field without components of the cluster, and a third intermediate region, which is out of analysis (void area), which
would correspond to a possible transition zone between the other two.

The correct choice of these radii, even having a previous estimation for the 'c+f' region, significantly affects the execution of the algorithm and,
in general, requires a generously wide field 'f'. There is no general rule that defines relative proportions of these areas.

Finally, when an acceptable result is obtained, what Clusterix provides is a probability field associated with the set of stars under study.
The dataset, with an extra column for the probability value of each object to belong to the open cluster, can be exported to TOPCAT
for further optimization and final processing. Figure \ref{fig:clusterix_probability} shows an example of the probability distribution in the proper
motion space for NGC 2682.

@@

Finally, it is necessary to refer to the work of \cite{castro2020hunting}, which presents a method based on machine learning techniques to make
a systematic search for overdensities in the astrometric space of the galactic disk and a subsequent identification of OC using photometric
information, also from Gaia DR2.

The method includes two phases: the first one uses an unsupervised clustering algorithm, DBSCAN, to search for overdensities
$(l, b \pi, \mu_{\alpha} *, \mu_{\delta})$, ans subsequently applies a deep learning ANN trained with magnitude diagrams to identify isochrone
patterns within the detected overdensities and thus proceed to confirm them as OC.

It should be noted that for the execution of this method, MareNostrum 42 (Barcelona Supercomputing Center) was used.
So the neural network could handle the image recognition process with isochrone patterns and not applying theoretical models
derived from values such as metallicity, masses, among other.

@@

There are several clustering algorithms: KMeans, Mean-Shift Clustering, DBSCAN \cite{ester1996density} among other.
While each one behaves better according ot the distribution of the objects to clusterize, we have choosen KMeans by its simplicity and good results.

@@

This algorithm works well and gives good results at first approach.
However, we need to set a large number of clusters to find the open cluster we are looking for.
This fact complicates the identification of the OC and many times the OC contains to many outliers.
For that reason, we have search for a KMeans refinement based on an Artificial Neural Network (ANN).

>>>>

An initial approach for finding OCs is the search for overdensities along the galactic disk.
In general, this is a good starting point but, although it seems simple,
it presents a fundamental problem already discussed.
The near field around the OC is filled with two types of distinct star populations:
those who belong to the OC (tens or hundreds to a few thousand)
and a background made up of thousands or millions of stars that are not.
Finding out which stars belong to the first group is the problem faced in this work.
This selection is crucial to properly characterize the fundamental properties of the cluster
(dynamics, total mass, age, chemical composition, etc.).

@@

However, as shown in Figures \ref{fig:pm_ngc_6494} and \ref{fig:parallax_ngc_6494},
in general it is not as easy and becomes necessary to consider other parameters such as distances,
or even metallicity and age (derived from isochrone curves).
Sometimes even, photometric data may be required for the stars within the studied field.

@@

At this point, it seems reasonable that the study of each individual group requires its own parametrization and technique
to identify it. With Gaia DR2, which contains high-quality astrometric and photometric data for a huge number of stars,
a new opportunity arises to simplify and optimize these processes. In this context, new approaches are being developed
to improve current detection methods and automate OC characterization.

@@

In this work we make use of Gaia DR2 since DR3 has not been released in time for us to include it.
Gaia DR2 is a multidimensional dataset obtained by ESA's Gaia mission (located at L2, 1.5 million kilometers from Earth) and
operational since 2014. The catalogue has high precision and accuracy astrometric data for more than 1.7 billion stellar sources,
and magnitudes in three photometric filters (G, BP and RP) for more than 1,300 million sources.
80\% of Gaia DR2 sources are weaker than $G_{mag} \approx 18$.
For magnitudes $G \approx 21$ nominal uncertainties reach $2 mas$ for parallaxes and $5 mas \cdot years^{-1}$ in proper motions.
But for closer sources, (at the bright end, $G < 14$), the precision is of the order of $0.02 mas$
and $0.05 mas \cdot year^{-1}$ respectively \cite{cantat2018gaia}.
In general, open clusters, which are the main subject of this work, are at optimal data thresholds.

In this context, and taking advantage of the opportunity provided by Gaia DR2,
new approaches are being developed to improve current detection methods and automating OC characterization as much as possible.
This is the case of tools such as Clusterix 2.0 \cite{balaguer2020clusterix}, TOPCAT \cite{taylor2005topcat}
and VOSA \cite{bayo2008vosa}, (all of them developed by the Virtual Observatory).
These tools work together and offer good results,
although it is not possible to use them in an unsupervised process and they require different parameterizations for each study case.

@@

TOPCAT by itself does not offer an independent discrimination mechanisms to researchers.
They have to try an initial selection, starting from the representation of proper motions,
by delimiting the area where the cluster most probably appears.
When this is not possible, it is necessary to have a previous knowledge about the cluster profile
and to parametrize some approximated values.
In both cases, the obtained selection must then be optimized step by step
until a result that meets the confidence levels is achieved.

Another tool is Clusterix 2.0, which is an interactive web-based tool \cite{balaguer2020clusterix}.
It takes the proper motion diagram without making any prior assumption about the membership of the candidate star
and determines empirically the frequency functions.

@@

$\left( a, b \right)$ are referred to the proper motion configuration space,
while a point located in the center of the cell $\left( i, j \right)$ provides the maximum
contribution for calculating the local density.
Number $h$ is called the \emph{smoothing} parameter and it is measured in the same units as proper motion.

Various clustering algorithms have been studied in the past with the purpose of finding a valid non-parameterized method.
UPMASK \cite{krone2014upmask} was developed to only use photometry and positions,
and then it was adapted to the Gaia data \cite{cantat2018gaia} based only on proper motion and parallax.
A similar approach has been carried out using DBSCAN and machine learning algorithms to discover new clusters.

In this sense, Clusterix also claims to be a non-parameterized method,
but it critically depends on the initial selection for the field sizes to be analyzed.

@@

This fact implies defining three areas or regions with different radius.
The first 'c+f' corresponds to the one in which the cluster members are
presumed to be contained together with other star fields that are not part of the cluster.
The second region is the broadest and assumes that it only contains stars
in an extended visual field without components of the cluster.
The third region is the intermediate one and is out of analysis (void area),
since it would correspond to a possible transition zone between the other two.

The right choice of these radii, even having a previous estimation for the 'c+f' region,
highly affects the execution of the algorithm and, in general, requires a considerable wide field 'f'.
There is no rule of thumb that defines relative proportions of these areas.

Finally, when an acceptable result is obtained,
what Clusterix provides is a probability field associated to the set of stars under study.
The recovered dataset contains all fields from Gaia database
and an extra column with the probability value of each object to belong to the open cluster.
This new dataset can be exported to TOPCAT for further optimization and final processing.
Figure \ref{fig:clusterix_probability} shows an example of the probability distribution
in the proper motion space for NGC 2682.

@@

Finally, it is necessary to refer to the work developed on \cite{castro2020hunting}.
It presents a method based on machine learning techniques to make a systematic search
for overdensities in the astrometric space of the galactic disk
and a later identification of OC using photometric information, also from Gaia DR2.

The method includes two phases: the first one uses an unsupervised clustering algorithm, DBSCAN,
to search for overdensities $(l, b \pi, \mu_{\alpha} *, \mu_{\delta})$,
and then applies a deep learning Artificial Neural Network (ANN),
previously trained with magnitude diagrams,
to identify isochrone patterns within the detected overdensities and thus proceed to confirm them as OC.

It should be noted that for the execution of this method,
MareNostrum 42 (Barcelona Supercomputing Center) was used.
So the neural network could handle the image recognition process with isochrone patterns
and not applying theoretical models derived from values such as metallicity or masses, among other.

@@

There are several clustering algorithms: KMeans, Mean-Shift Clustering, DBSCAN \cite{ester1996density} among other.
While each one behaves better according to the distribution of the objects to clusterize,
we have chosen KMeans by its simplicity and good results.

@@

This algorithm works well and gives good results at first approach.
However, we need to set a large number of clusters to find the open cluster we are looking for.
This fact complicates the identification of the OC and many times the found cluster contains to many outliers.
For that reason, we have search for a KMeans refinement based on an artificial neural network.
